---
title: "Data Cleaning Report"
subtitle: "Generated using [cuci](https://folkehelsestats.github.io/cuci)) R package"
date: "`r format(Sys.time(), '%d %B %Y, %H:%:%M')`"
author: "`r Sys.info()[['user']]`"


format:
  html:
    theme: cosmo

    # Layout
    page-layout: full

    # Wider content area
    grid:
      sidebar-width: 220px
      body-width: 1600px
      margin-width: 50px
      gutter-width: 1rem

    # TOC
    toc: true
    toc-depth: 3
    toc-location: left
    toc-title: "Contents"

    # Code
    code-fold: true
    code-summary: "Show code"

    # HTML options
    self-contained: true
    smooth-scroll: true
    highlight-style: github

    # Custom styles
    css: audit_report.css

execute:
  echo: false
  warning: false
  message: false

params:
  log_dir: "logs/matching"
---

```{r setup}
library(data.table)
library(DT)           # interactive tables
library(ggplot2)      # charts
library(scales)       # formatting helpers

# ── Load master log ────────────────────────────────────────────────
master_path <- file.path("O:\\Prosjekt\\Rusdata\\Rusundersokelsen\\Datasets\\temp\\logs\\match_log_MASTER.csv")
# master_path <- file.path(params$log_dir, "match_log_MASTER.csv")

if (!file.exists(master_path)) {
  stop(sprintf(
    "Master log not found at '%s'. Run the cleaning pipeline first.", master_path
  ))
}

master <- fread(master_path)

# ── Derived fields used throughout the report ─────────────────────

# Confidence as an ordered factor for consistent sorting/colouring
master[, confidence := factor(
  confidence,
  levels = c("high", "medium", "low", "none"),
  ordered = TRUE
)]

# Traffic-light badge labels
master[, badge := fcase(
  method == "exact",           "Exact",
  method == "fuzzy",           "Fuzzy",
  grepl("^keyword", method),   "Predicted",
  method == "unmatched",       "Unmatched",
  default = "Unknown"
)]

# Counts used in summary boxes
n_total     <- nrow(master)
n_exact     <- master[badge == "Exact",     .N]
n_fuzzy     <- master[badge == "Fuzzy",     .N]
n_predicted <- master[badge == "Predicted", .N]
n_unmatched <- master[badge == "Unmatched", .N]
n_review    <- master[needs_review == TRUE, .N]
n_datasets  <- master[, uniqueN(dataset)]
```

```{r helpers}
# ── Inline HTML badge helper ───────────────────────────────────────
badge_html <- function(label) {
  colour <- switch(label,
    "Exact"     = "#2e7d32",   # green
    "Fuzzy"     = "#e65100",   # amber
    "Predicted" = "#b71c1c",   # red
    "Unmatched" = "#424242",   # grey
    "#555"
  )
  sprintf(
    '<span style="background:%s;color:#fff;padding:2px 8px;
     border-radius:4px;font-size:0.78em;font-weight:600;
     letter-spacing:0.04em">%s</span>',
    colour, label
  )
}

# Apply badge HTML to a vector (used in DT tables)
badge_vec <- Vectorize(badge_html)
```

## Overview

::: {.callout-note icon=false}
This report summarises every column mapping decision made during the
automated data cleaning pipeline. Rows flagged **Fuzzy** or **Predicted**
require manual verification before the cleaned data is used for analysis.
:::

### Summary

```{r summary-boxes}
#| echo: false

# Render four KPI cards as inline HTML
cards <- list(
  list(val = n_datasets,  lab = "Datasets processed",   col = "#1565c0"),
  list(val = n_exact,     lab = "Exact matches",         col = "#2e7d32"),
  list(val = n_fuzzy,     lab = "Fuzzy matches ⚠",      col = "#e65100"),
  list(val = n_predicted, lab = "Predicted matches ⚠⚠", col = "#b71c1c")
)

htmltools::div(
  class = "kpi-row",
  lapply(cards, function(c) {
    htmltools::div(
      class = "kpi-box",
      style = sprintf("border-top:4px solid %s", c$col),

      htmltools::div(
        class = "kpi-val",
        style = sprintf("color:%s", c$col),
        c$val
      ),

      htmltools::div(
        class = "kpi-lab",
        c$lab
      )
    )
  })
)
```

```{r review-alert}
#| output: asis

if (n_review > 0) {
  cat(sprintf(
    '::: {.callout-warning}\n**%d column mapping(s) require your review** 
    (fuzzy + predicted matches combined). 
    See the [Review Checklist](#review-checklist) section below.\n:::\n',
    n_review
  ))
} else {
  cat('::: {.callout-tip}\n✔ All column matches were exact — no manual review needed.\n:::\n')
}
```

---

### Match Method Breakdown

```{r method-chart, fig.height=3.2, fig.width=7}
method_counts <- master[, .N, by = badge]
method_counts[, badge := factor(badge, levels = c("Exact","Fuzzy","Predicted","Unmatched"))]

pal <- c(
  "Exact"     = "#2e7d32",
  "Fuzzy"     = "#e65100",
  "Predicted" = "#b71c1c",
  "Unmatched" = "#9e9e9e"
)

ggplot(method_counts, aes(x = badge, y = N, fill = badge)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = N), vjust = -0.4, fontface = "bold", size = 4.2) +
  scale_fill_manual(values = pal) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = "Number of columns") +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.x        = element_text(face = "bold")
  )
```

---

### Per-Dataset Summary

```{r per-dataset-table}
per_ds <- master[, .(
  Total    = .N,
  Exact    = sum(badge == "Exact"),
  Fuzzy    = sum(badge == "Fuzzy"),
  Predicted= sum(badge == "Predicted"),
  Unmatched= sum(badge == "Unmatched"),
  `Needs Review` = sum(needs_review)
), by = .(Dataset = dataset, Year = year)][order(Year)]

datatable(
  per_ds,
  rownames  = FALSE,
  options   = list(dom = "t", pageLength = 20, ordering = TRUE),
  class     = "stripe hover compact"
) |>
  formatStyle("Needs Review",
    color            = styleInterval(0, c("inherit", "white")),
    backgroundColor  = styleInterval(0, c("inherit", "#b71c1c")),
    fontWeight       = "bold"
  ) |>
  formatStyle("Exact",
    color = styleInterval(0, c("inherit", "#2e7d32")),
    fontWeight = "bold"
  )
```

---

## Review Checklist {#review-checklist}

Columns below were mapped by **fuzzy matching** (possible typos) or
**keyword prediction** (semantic guessing). Tick off each one after
verifying it is correct. If a mapping is wrong, add the correct alias
to `config/variable_map.yaml` and re-run the pipeline.

```{r review-table}
review_dt <- master[needs_review == TRUE][order(confidence)]

if (nrow(review_dt) == 0) {
  cat("_No columns flagged for review._\n")
} else {
  display_dt <- review_dt[, .(
    Dataset   = dataset,
    Year      = year,
    `Raw column`      = raw_name,
    `Mapped to`       = canonical,
    Method    = method,
    Type      = badge,
    Action    = action
  )]

  # Inject coloured badge HTML into the Type column
  display_dt[, Type := badge_vec(Type)]

  datatable(
    display_dt,
    escape    = FALSE,          # allow HTML badges to render
    rownames  = FALSE,
    filter    = "top",          # per-column filter boxes
    options   = list(
      pageLength = 15,
      dom        = "Bfrtip",
      buttons    = c("csv", "excel"),
      columnDefs = list(
        list(className = "dt-center", targets = c(1, 5))
      )
    ),
    extensions = "Buttons",
    class      = "stripe hover compact"
  )
}
```

---

## Full Matching Log {#full-log}

Every column from every dataset — exact matches included. Use the
filters and search box to explore.

```{r full-log-table}

full_display <- master[, .(
  Dataset        = dataset,
  Year           = year,
  `Raw column`   = raw_name,
  `Canonical`    = canonical,
  Method         = method,
  Confidence     = as.character(confidence),
  `Needs review` = needs_review,
  Type           = badge,
  Action         = action,
  `Logged at`    = logged_at
)]

full_display[, Type := badge_vec(Type)]

datatable(
  full_display,
  escape     = FALSE,
  rownames   = FALSE,
  filter     = "top",
  options    = list(
    pageLength = 20,
    dom        = "Bfrtip",
    buttons    = c("csv", "excel"),
    scrollX    = TRUE
  ),
  extensions = "Buttons",
  class      = "stripe hover compact"
) |>
  formatStyle("Needs review",
    color           = styleEqual(c(TRUE, FALSE), c("white",  "inherit")),
    backgroundColor = styleEqual(c(TRUE, FALSE), c("#b71c1c","inherit")),
    fontWeight      = "bold"
  )
```

---

## Unmatched Columns {#unmatched}

These columns could not be mapped to any canonical variable and were
**dropped** from the clean dataset. If they should be retained, add
them to `config/variable_map.yaml`.

```{r unmatched-table}
unmatched_dt <- master[badge == "Unmatched"]

if (nrow(unmatched_dt) == 0) {
  cat("_No unmatched columns — all raw columns were successfully mapped._\n")
} else {
  datatable(
    unmatched_dt[, .(Dataset = dataset, Year = year, `Raw column` = raw_name)],
    rownames = FALSE,
    options  = list(dom = "frtip", pageLength = 20),
    class    = "stripe hover compact"
  )
}
```

---

## Pipeline Configuration

```{r config-info}
#| output: asis
cat(sprintf("- **Log directory:** `%s`\n", params$log_dir))
cat(sprintf("- **Master log:** `%s`\n", master_path))
cat(sprintf("- **Report generated:** %s\n", format(Sys.time(), "%Y-%m-%d %H:%M:%S")))
cat(sprintf("- **R version:** %s\n", R.version.string))
cat(sprintf("- **data.table version:** %s\n", packageVersion("data.table")))
```
