---
title: "Introduction to cuci"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to cuci}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

## What cuci does

Survey datasets collected over multiple years rarely arrive with identical
column names, value codings, or file formats. **cuci** (Clean and Unify
Collected Information) solves this by letting you declare — once, in a YAML
file — what every variable should look like: its canonical name, expected
column aliases, target data type, valid values, recode rules, and missing-value
codes. The package then applies that declaration consistently to every raw file
you give it.

The core idea is that **the YAML is the specification and R is the executor**.
You spend your time editing the YAML; cuci does the mechanical work.

---

## Package architecture

cuci is built as a chain of small, single-purpose functions. Understanding the
chain makes it easy to know which function to call when you do not want to run
the whole pipeline.

```
load_config()          Read and parse variable_map.yml into R
      |
      v
match_columns()        Map raw column names to canonical names (3 layers)
      |
      v
clean_dataset()        Recode, coerce, validate one dataset
      |
      v
validate_dataset()     Print a post-clean quality report
      |
      v
export_match_log()     Write audit CSVs for the decisions just made
      |
      v
run_pipeline()         Orchestrates all of the above for a list of files
```

Every step in this chain is also a standalone exported function you can call
directly. The sections below show both the full pipeline and each function on
its own.

---

## 1. The variable map YAML

Everything starts with a YAML file that describes your target dataset. Each
top-level key is the canonical name of a variable — the name it will have in
the cleaned output, regardless of what it was called in the raw file.

```yaml
# variable_map.yml

kjonn:
  colnames: [Kjonn, IO_Kjonn, gender_var]    # known aliases in raw files
  keywords: [kjonn, sex, gender]             # fallback keywords for prediction
  type: integer                              # target R type after cleaning
  label: "Respondentens kjønn"
  value:                                     # complete set of valid values
    1: "Mann"
    2: "Kvinne"
  recode:                                    # map old codes -> standard codes
    "0": 2
    "1": 1
  missing:                                   # codes that mean "no answer"
    - 8
    - 9

alder:
  colnames: [Alder, age_yrs]
  keywords: [alder, age, ageyrs]
  type: integer
  label: "Alder i år"
  value: ~                                   # ~ means no value constraint
  recode: ~
  missing:
    - NA
    - 99
```

**`colnames`** lists every name this variable might have in a raw file.
Matching against this list is always tried first (exact match), and is the
most reliable layer.

**`keywords`** are used as a fallback when no exact match is found. The
package builds a regex from these words and tests it against unrecognised
column names. A keyword match is always flagged for human review because it
is a prediction, not a certainty.

**`type`** is the R type the column should have after cleaning. If a column
cannot be safely converted (e.g. it contains `"D"` and the target type is
`integer`) the column is left unchanged and the problem is recorded in the
issues table — no silent data loss.

**`value`** declares the complete set of valid integer codes. Any code found
in the data that is not in this list triggers a warning and is recorded as an
issue. The values are not removed; that decision belongs to the analyst.

**`recode`** maps old codes to standard ones before type coercion runs. Keys
are always strings (matching the raw string representation of the cell),
values are integers. Use this when older survey versions used `0/1/2` where
the standard is `1/2`.

**`missing`** lists codes that should be recoded to `NA`. This step is
**opt-in** (`apply_missing = TRUE`) because silently turning plausible
integers like `8` or `9` into `NA` can hide real problems. Declare the
missing codes in the YAML so the intent is documented, and decide at
pipeline-run time whether to apply them.

---

## 2. Full pipeline — `run_pipeline()`

`run_pipeline()` is the entry point for routine processing. Give it a
manifest table (one row per file) and it reads, matches, cleans, validates,
and logs everything in sequence.

```{r}
library(cuci)
library(data.table)

# Build a manifest: one row per raw survey file
manifest <- data.table(
  path     = c(
    "data/raw/survey_2022.csv",
    "data/raw/survey_2023.csv",
    "data/raw/survey_2024.csv"
  ),
  year_tag = c(2022L, 2023L, 2024L)
)

# merge = FALSE (default): returns a named list, one data.table per file
results <- run_pipeline(
  file_manifest = manifest,
  config_file   = "config/variable_map.yml",
  log_dir       = "logs/matching",
  output_path   = "data/clean",       # directory: writes one CSV per file
  apply_missing = FALSE,
  merge         = FALSE
)

# Access individual years
results[["survey_2022.csv"]]
results[["survey_2023.csv"]]
```

### Merging into one master dataset

When you want all years stacked into a single table:

```{r}
master <- run_pipeline(
  file_manifest = manifest,
  config_file   = "config/variable_map.yml",
  log_dir       = "logs/matching",
  output_path   = "data/clean/master.csv",  # file path, not directory
  merge         = TRUE
)

# master is a single data.table with a 'year' column as the first column
head(master)
```

Columns that exist in some years but not others are padded with `NA` so the
stack always works. `year` is moved to the first column automatically.

---

## 3. Using functions individually

`run_pipeline()` is convenient for routine work, but every step is also an
exported function you can call on its own. You would do this when:

- **Exploring a new dataset** before committing to a full pipeline run.
- **Debugging** a specific step without re-running the whole chain.
- **Building a custom workflow** that needs only some of the steps (e.g. you
  already have cleaned data but want the audit log, or you want to check
  column matching without writing any files).
- **Testing a new variable_map.yml** entry before rolling it out to all files.

The sections below show each function individually and explain when reaching
for it directly makes sense.

---

### 3.1 `load_config()` — parse the YAML once

```{r}
config <- load_config("config/variable_map.yml")
```

`load_config()` reads the YAML file and builds seven lookup structures from
it. The result is a plain R list you can inspect interactively at any time.

```{r}
# What variables are defined?
names(config$var_map)
#> [1] "kjonn"       "alder"       "yrkesstatus" "siv"         "antpers"
#> [6] "tob1"

# What type is each variable expected to be?
config$type_map
#>       kjonn       alder yrkesstatus         siv     antpers        tob1
#>   "integer"   "integer"   "integer"   "integer"   "integer"   "integer"

# What are the valid values for kjonn?
config$value_map$kjonn
#>    num_value chr_value
#> 1:         1      Mann
#> 2:         2    Kvinne

# What are the recode rules for kjonn?
config$recode_map$kjonn
#>    raw_value new_value
#> 1:         0         2
#> 2:         1         1
#> 3:         2         2

# What column aliases are recognised for kjonn?
names(config$name_lookup)[config$name_lookup == "kjonn"]
#> [1] "Kjonn"    "kjonn"    "IO_Kjonn" "io_kjonn"
```

**When to call directly:** every time you want to examine or verify your
variable map. Also useful in testing: load the config once and pass it to
multiple test cases without re-reading the YAML each time.

---

### 3.2 `match_columns()` — test column name matching

`match_columns()` takes a character vector of column names (as they appear in
a raw file after normalisation) and returns a matching decision for each one.

```{r}
library(data.table)

# Read a raw file and normalise its column names
raw <- fread("data/raw/survey_2023.csv")

# See what names the file actually has
names(raw)
#> [1] "IO_Kjonn"    "Alder"       "Yrkesstatus" "Siv"         "Antpers"
#> [6] "Tob1"

# Run matching directly to inspect decisions before cleaning
result <- match_columns(
  raw_colnames = tolower(names(raw)),   # simulate normalisation
  config       = config
)

# What was matched, and how?
result$match_log
#>       raw_name   canonical           method confidence needs_review
#> 1:   io_kjonn       kjonn  keyword [kjonn]        low         TRUE
#> 2:       alder       alder            exact       high        FALSE
#> 3: yrkesstatus yrkesstatus            exact       high        FALSE
#> 4:         siv         siv            exact       high        FALSE
#> 5:     antpers     antpers            exact       high        FALSE
#> 6:        tob1        tob1            exact       high        FALSE

# What was the rename map? (old name -> canonical name)
result$rename_vec
#>  io_kjonn
#>   "kjonn"

# What could not be matched at all?
result$unmatched
#> character(0)
```

The match is done in three layers:

1. **Exact** — the column name is a known alias in the YAML `colnames:` list.
   Confidence: high. No review needed.
2. **Fuzzy** — the column name is close to a known alias by edit distance
   (catches typos). Confidence: medium. Flagged for review.
3. **Keyword** — the column name contains one of the `keywords:` from the
   YAML. Confidence: low. Always flagged for review because it is a
   prediction, not a guaranteed match.

**When to call directly:** whenever you receive a new file with unfamiliar
column names and want to see what the package will do before committing to a
full run. The `match_log` and `print_match_report()` output tell you
immediately which columns need attention in the YAML.

```{r}
# Pretty-print the matching decisions to the console
print_match_report(result, dataset_label = "survey_2023.csv")
```

---

### 3.3 `clean_dataset()` — clean one dataset

`clean_dataset()` takes a raw `data.table` and a config, and returns a list
with two elements: the cleaned data and an issues table.

```{r}
raw <- fread("data/raw/survey_2023.csv")

out <- clean_dataset(
  dt            = raw,
  config        = config,
  year_tag      = 2023L,
  dataset_label = "survey_2023.csv",
  apply_missing = FALSE              # keep 8/9 as integers for now
)

# The cleaned data.table
out$data
#>     year kjonn alder yrkesstatus siv antpers tob1
#>  1: 2023     2    75           2   3       1    2
#>  ...

# The issues table — one row per problem found
out$issues
#>    variable    issue_type                                  detail
#> 1:      siv coerce_failure  Target: integer. 1 value(s) -> NA: D...
#> 2:     tob1 unexpected_values  Values not declared in YAML: 8, 9...
```

`clean_dataset()` never silently corrupts data. If a column cannot be
converted to its declared type without introducing new `NA` values, it is
left at its original type and the problem is recorded in `out$issues`. Check
this table before saving the cleaned data.

**When to call directly:** when processing a single file in an exploratory or
interactive session, or when you want to inspect the issues before deciding
whether to write the output. The issues table is the primary tool for
understanding what the YAML needs to be updated to handle.

```{r}
# Workflow: clean interactively, inspect issues, update YAML, repeat
out <- clean_dataset(raw, config, year_tag = 2023L, dataset_label = "survey_2023.csv")

if (nrow(out$issues) > 0) {
  print(out$issues)
  # Fix variable_map.yml, reload config, try again
}
```

---

### 3.4 `validate_dataset()` — check the cleaned result

`validate_dataset()` prints a structured report about a cleaned dataset:
row count, column names, missingness per column, and any values that are
present in the data but not declared in the YAML `value:` blocks.

```{r}
validate_dataset(out$data, config, dataset_label = "survey_2023.csv")
#>
#> ================================================
#>  Validation Report: survey_2023.csv
#> ================================================
#>   Rows    : 20
#>   Columns : year, kjonn, alder, yrkesstatus, siv, antpers, tob1
#>
#>   Missingness per column: (<-- HIGH indicates >50% missing)
#>     year                 0%
#>     kjonn                0%
#>     alder                0%
#>     yrkesstatus          0%
#>     siv                  5%
#>     antpers              0%
#>     tob1                 0%
#>
#>   Value conformance issues:
#>     ⚠ tob1                 unexpected: 8, 9           valid: 1, 2
#> ================================================
```

The function returns the data invisibly, so it can be placed inline in a
pipeline without breaking the chain.

```{r}
# Inline use: validate and immediately assign
cleaned <- clean_dataset(raw, config, year_tag = 2023L) |>
  (\(x) { validate_dataset(x$data, config); x$data })()
```

**When to call directly:** after `clean_dataset()` when you want a summary
before saving. Also useful when loading an already-cleaned CSV from disk and
want to verify it still conforms to the current YAML (e.g. after updating
the variable map).

```{r}
# Check an already-cleaned file against the current YAML
existing <- fread("data/clean/survey_2023_clean.csv")
validate_dataset(existing, config, "survey_2023_clean.csv")
```

---

### 3.5 `export_match_log()` — write audit logs without running the pipeline

`export_match_log()` writes three CSV files from the result of
`match_columns()`:

- `match_log_<dataset>.csv` — every column from the raw file, what it was
  matched to, by which method, and whether it needs review.
- `issues_<dataset>.csv` — coercion failures and unexpected values from
  `clean_dataset()`. Only written if there are issues.
- `match_log_MASTER.csv` — a single file that accumulates rows from all
  datasets, updated on every call. Re-processing the same dataset replaces
  its old rows rather than duplicating them.

```{r}
raw <- fread("data/raw/survey_2023.csv")

# Step 1: match column names
match_result <- match_columns(
  raw_colnames = tolower(names(raw)),
  config       = config
)

# Step 2: clean (to get the issues table)
out <- clean_dataset(raw, config, year_tag = 2023L, dataset_label = "survey_2023.csv")

# Step 3: write all log files
export_match_log(
  match_result  = match_result,
  issues_dt     = out$issues,
  dataset_label = "survey_2023.csv",
  year_tag      = 2023L,
  log_dir       = "logs/matching",
  append_master = TRUE
)
#> Log directory: C:/Projects/survey/logs/matching
#>   Match log saved: C:/Projects/survey/logs/matching/match_log_survey_2023_csv.csv
#>   Issues log saved: C:/Projects/survey/logs/matching/issues_survey_2023_csv.csv
#>   Master log updated: C:/Projects/survey/logs/matching/match_log_MASTER.csv (6 total decisions)
```

**When to call directly:** when you have already cleaned a dataset
interactively and now want to produce the audit trail for it without
re-running the full pipeline. Also useful when you want to log matching
decisions from an exploratory `match_columns()` call — just pass
`issues_dt = NULL` to skip the issues log.

```{r}
# Log matching decisions only, no issues table
export_match_log(
  match_result  = match_result,
  issues_dt     = NULL,             # skip issues log
  dataset_label = "survey_2023.csv",
  year_tag      = 2023L,
  log_dir       = "logs/matching",
  append_master = FALSE             # do not update the master log yet
)
```

Setting `append_master = FALSE` is useful when you are iterating on a new
dataset and do not want half-finished entries accumulating in the master log
until you are satisfied with the result.

---

### 3.6 `summarise_master_log()` — review all decisions at once

After processing several datasets, call `summarise_master_log()` to get a
cross-dataset overview of every matching decision recorded in
`match_log_MASTER.csv`.

```{r}
summarise_master_log("logs/matching")
#>
#> ╔══════════════════════════════════════════════════════╗
#>   MASTER AUDIT SUMMARY — All Datasets
#> ╚══════════════════════════════════════════════════════╝
#>
#>   Decisions by match method (all datasets):
#>     exact                          268
#>     unmatched                      2
#>     keyword [kjonn]                1
#>
#>   Columns flagged for review per dataset:
#>     survey_2022.csv                     (year 2022): 1 column(s)
#>     survey_2023.csv                     (year 2023): 1 column(s)
#>
#>   ⚠⚠ ALL PREDICTED (keyword) MATCHES — verify these:
#>     [survey_2022.csv | 2022]  io_kjonn  -> kjonn  (keyword [kjonn])
#>     [survey_2023.csv | 2023]  io_kjonn  -> kjonn  (keyword [kjonn])
```

The function returns the full master log `data.table` invisibly so you can
work with it programmatically.

```{r}
# Load the master log for custom analysis
master_log <- summarise_master_log("logs/matching")

# Find all columns that were dropped (unmatched) across all datasets
dropped <- master_log[method == "unmatched"]
dropped[, .(dataset, year, raw_name)]
```

**When to call directly:** at the end of any processing session to confirm
that all keyword-predicted matches are sensible, and that no unexpected
columns were silently dropped.

---

## 4. Typical workflows

### Workflow A — exploring a new dataset

You have received a new CSV file and want to understand how it maps to your
standard variable set before running anything that writes files.

```{r}
library(cuci)
library(data.table)

config <- load_config("config/variable_map.yml")
raw    <- fread("data/raw/new_survey.csv")

# Step 1: see what columns the file has and how they match
match_result <- match_columns(tolower(names(raw)), config)
print_match_report(match_result, "new_survey.csv")

# Step 2: if the match looks wrong, update variable_map.yml and reload
# config <- load_config("config/variable_map.yml")

# Step 3: clean and inspect issues
out <- clean_dataset(raw, config, year_tag = 2025L, dataset_label = "new_survey.csv")
out$issues

# Step 4: validate the result
validate_dataset(out$data, config, "new_survey.csv")
```

At no point in this workflow is anything written to disk. You are purely
inspecting. Once you are satisfied, you can either call `export_match_log()`
to save the audit trail, or add the file to the manifest and run
`run_pipeline()`.

---

### Workflow B — fixing a specific variable

You have noticed that `tob1` is producing unexpected-value warnings across
all years. You want to understand the problem without re-processing everything.

```{r}
config <- load_config("config/variable_map.yml")

# Inspect the current declaration for tob1
config$value_map$tob1      # what values are currently declared valid
config$missing_map$tob1    # what values are declared as missing codes

# Load one problem file and look at the raw tob1 column
raw <- fread("data/raw/survey_2023.csv")
table(raw$Tob1, useNA = "always")
#>  1  2  8  9 <NA>
#> 14  4  1  1    0

# 8 and 9 are present — they are the "no answer" / "don't know" codes
# They are declared in missing: but apply_missing was FALSE
# Solution: either add them to value: (to acknowledge they appear)
# or re-run with apply_missing = TRUE to convert them to NA

out_with_missing <- clean_dataset(
  raw, config,
  year_tag      = 2023L,
  dataset_label = "survey_2023.csv",
  apply_missing = TRUE    # now 8 and 9 become NA before value check
)
out_with_missing$issues   # tob1 should no longer appear here
```

---

### Workflow C — adding a new variable to the YAML

You want to add a new variable `utdanning` to the pipeline. You can test the
new YAML entry against a single file before running the full manifest.

```{r}
# Edit variable_map.yml to add utdanning, then:
config <- load_config("config/variable_map.yml")

# Verify the new variable is present
"utdanning" %in% names(config$var_map)
config$type_map[["utdanning"]]
config$value_map$utdanning

# Check that it matches correctly in one file
raw          <- fread("data/raw/survey_2023.csv")
match_result <- match_columns(tolower(names(raw)), config)
match_result$match_log[canonical == "utdanning"]

# If it does not match, check the aliases and keywords
config$name_lookup[config$name_lookup == "utdanning"]
```

---

### Workflow D — re-running a single file after a YAML fix

You fixed a problem in `variable_map.yml` for one year's file and want to
re-generate its log without re-processing every other year.

```{r}
config <- load_config("config/variable_map.yml")
raw    <- fread("data/raw/survey_2022.csv")

out <- clean_dataset(raw, config, year_tag = 2022L, dataset_label = "survey_2022.csv")
validate_dataset(out$data, config, "survey_2022.csv")

# Write the log — .update_master_log() will replace the old 2022 rows
export_match_log(
  match_result  = match_columns(tolower(names(raw)), config),
  issues_dt     = out$issues,
  dataset_label = "survey_2022.csv",
  year_tag      = 2022L,
  log_dir       = "logs/matching",
  append_master = TRUE    # replaces old rows for survey_2022.csv in the master
)
```

The master log uses `dataset_label` to identify which rows belong to each
file. When `append_master = TRUE`, any existing rows for `"survey_2022.csv"`
are deleted before the new rows are inserted, so there is no duplication.

---

## 5. The audit log files

Three CSV files are produced in `log_dir` over the life of a project.

**`match_log_<dataset>.csv`** — one per dataset, overwritten on each run. Contains one row per column from the raw file. The most useful columns for a human reviewer are:

| Column | Meaning |
|---|---|
| `raw_name` | The column name as it appeared in the raw file (after normalisation) |
| `canonical` | The canonical name it was mapped to (`NA` if unmatched) |
| `method` | `exact`, `fuzzy`, `keyword [word]`, or `unmatched` |
| `confidence` | `high`, `medium`, `low`, or `none` |
| `needs_review` | `TRUE` for fuzzy, keyword, and unmatched rows |
| `action` | Plain-English description of what was done |

**`issues_<dataset>.csv`** — one per dataset, only written when there are problems. Contains one row per coercion failure or unexpected value.

| Column | Meaning |
|---|---|
| `variable` | The canonical variable name |
| `issue_type` | `coercion_failure`, `coercion_skip`, or `unexpected_values` |
| `detail` | Description of the problem, including the bad values |

**`match_log_MASTER.csv`** — accumulates rows from all datasets. The primary tool for cross-year auditing. Open this file in Excel or load it with `fread()` to see all matching decisions across the entire project in one place.

```{r}
# Load and filter the master log programmatically
master <- data.table::fread("logs/matching/match_log_MASTER.csv", quote = "")

# All keyword-predicted matches (require human verification)
master[grepl("^keyword", method)]

# All unmatched columns across all datasets (were silently dropped)
master[method == "unmatched", .(dataset, year, raw_name)]

# All columns flagged for review, sorted by year
master[needs_review == TRUE][order(year)]
```
