---
title: "Rendering Audit Reports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering Audit Reports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview

The **cuci** package can generate HTML audit reports summarizing the
results of a column-matching workflow.

The report is created using the bundled Quarto template and the matching
log files produced during the workflow pipeline.

The main function to generate report is:

```{r, eval = FALSE}
cuci::render_report()
```

A successful rendering requires a log directory containing the master
matching log file that is created automatically during data cleaning with `run_pipeline()`:

```
match_log_MASTER.csv
```

# Requirements

Report rendering requires a working installation of Quarto.

The Quarto installation can be checked with:

```{r, eval = FALSE}
quarto::quarto_version()
```

If Quarto is not installed, download it from:

<https://quarto.org/docs/get-started/>

# Basic usage

Assuming matching logs are stored in:

```
logs/matching/
```

and the directory contains:

```
logs/matching/
└── match_log_MASTER.csv
```

a report can be generated with:

```{r, eval = FALSE}
library(cuci)

render_report(
  log_dir = "logs/matching"
)
```

The function returns the path to the generated HTML report.

```{r, eval = FALSE}
report_path <- cuci::render_report(
  log_dir = "logs/matching"
)

report_path
```

# Output location

By default, the report is saved in the user's `Documents` directory.

A custom output directory can be specified:

```{r, eval = FALSE}
render_report(
  log_dir = "logs/matching",
  output_dir = "reports"
)
```

If the directory does not exist, `render_report()` asks for
confirmation before creating it in an interactive session.

To create directories automatically:

```{r, eval = FALSE}
render_report(
  log_dir = "logs/matching",
  output_dir = "reports",
  create_dir = TRUE
)
```

# Custom filenames

The output filename can be specified using `output_file` else use the default file naming with current date:

```{r, eval = FALSE}
render_report(
  log_dir = "logs/matching",
  output_dir = "reports",
  output_file = "audit_report.html"
)
```

# Automated workflows

For scripts or automated pipelines, disable browser opening and avoid
interactive prompts:

```{r, eval = FALSE}
render_report(
  log_dir = "logs/matching",
  output_dir = "reports",
  output_file = "audit_report.html",
  create_dir = TRUE,
  open_browser = FALSE
)
```

